---
title: "B. Health Impact Analysis"
author: "Kevin Donkers (EI CDT) - 700063874"
date: "12/01/2021"
output:
  bookdown::html_document2:
    fig_caption: yes
    number_sections: false
    smart: no
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: false
    toc: false
---
```{r setup, include=FALSE}
library(here)
library(bookdown)
knitr::opts_chunk$set(echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE, include=FALSE)
# knitr::opts_chunk$set(echo=TRUE, message=TRUE, cache=FALSE, warning=FALSE, include=TRUE)
knitr::read_chunk(here::here('scripts/init.R'))
knitr::read_chunk(here::here('scripts/question1.R'))
knitr::read_chunk(here::here('scripts/question2.R'))
```


```{r load_data}
```

# Q1. Summarise the number of hospital admissions for COPD in England between 2001 and 2010.

```{r 1.init}
```

Data for hospital admissions for COPD in England are provided in the `copdobserved.csv` dataset. We can summarise the data as follows:

```{r 1.summary, include=TRUE}
```

From this we can see that for the whole of England over the ten years of data collected some local authorities reported zero cases, while the maximum number of cases in one local authority was 480 in a year. The country as a whole experienced an average of 67.0 annual cases of COPD per local authority.

To determine whether there was any overall trend in the total number of cases reported in England each year we can plot a line graph:

```{r 1.time.sum.data}
```
```{r 1.time.sum.plot, include=TRUE}
```

From this graph we can see that there is no overall trend in the total number of cases of COPD in England over the 2001-2010 period. Therefore we can determine that nationally the total number of cases are steady over time.

We can summarise the spatial spread of the data by plotting the mean number of cases for each local authority over the ten year time period onto a map:

```{r 1.eng.mean.data}
```
```{r 1.eng.mean.map, include=TRUE}
```

From this a notable few regions stand out as having particularly high average cases. These are the North East of England, the urban North of England, Birmingham and Cornwall. The high average cases in urban North of England could be explained by more exposure to air pollution than in more rural areas, as could the higher cases in the North East and Birmingham. Perhaps the (former) industrial nature of these regions plays a part as well. Cornwall is a more rural county, so higher cases here might be explained by the higher prevelance of radon gas in houses due to the local type of stone used to build them. 

To conclude, there is a relatively large spread in the number of cases of COPD per local authority in England, with some notable hotspots. There does not seems to be a time trend at a national level over the ten years of data.

# Q2. Estimate raw and smoothed standardised morbidity ratios (SMR). You should look for any spatial(!) patterns and/or extreme values and consider any potential reasons for them. You should note any differences between raw and the smoothed estimates. If there are big changes, are they appropriate?

Raw SMRs can be calculated by simply calculating the ratio between observed and expected admissions for COPD for each year and local authority. Unfortunately smoothed SMRs could not be calculate for this assessment because every time a model was initiated R crashed with a "fatal error". Given this (unhelpful) error message I was not able to debug and get the model to run.

Plotting raw SMR values on a map of England highlights a different spread of cases to that of the mean values mapped in question 1:

```{r 2.init}
```
```{r 2.raw.smrs}
```
```{r 2.map.smrs, include=TRUE}
```

We can see that conclusions from question 1 about urban and industrial areas (North East and Northern England) having higher COPD cases due to poorer air quality holds up. Conclusions about Cornwall having high COPD case numbers due to radon exposure now appear to be moot given the raw SMR values are around 1. Therefore the high average cases in Cornwall are likely due to demographic effects i.e. aged population.

We can pick out the top ten highest mean raw SMRs and display them in a table:

```{r 2.top.smrs, include=TRUE}
```

After a bit of research these locations are either large cities (e.g. Liverpool), part of large cities (e.g. Tower Hamlets) or industrial towns (e.g. Corby). It is notable also that many of these locations tend to be to the east of large cities, which tend to be down-wind of city centres given that prevailing winds tend to be westerly. Therefore poor air quality in the centre of cities can affect those living downwind as well as in city centres.



# Q3. Investigate any changes in the risks of hospitalisation due to COPD in England over time.




# Q4. Suppose a government minister wants to increase resources to hospitals to cope with admissions for COPD. Write a short report (maximum 1 page text) recommending key areas/local authorities in England where these resources should be allocated. In your answer you should explain the key steps of the analysis you have done, providing comment on what conclusions you can draw and any limitations associated. You do not need to include every step of the analyses, but should include results, figures and tables that support your findings. Ensure any plots and/or tables included are labelled clearly and are accompanied with appropriate captions.